---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setup}
  library(knitr)
  opts_knit$set(root.dir=normalizePath('./'))
  opts_chunk$set(fig.path = "./figures/") # corrected path
```

## Loading and preprocessing the data
```{r}
getData <- function() {
        partData <- read.table("./activity.csv", sep = ",", header = TRUE)
        #Add the date column
        partData$date <- as.Date(partData$date, format="%Y-%m-%d")
        #Add the day column
        partData$day <- weekdays(partData$date)
        #add the weekday column
        partData$weekday <- apply(partData, 1, function(d) {   
                if(d["day"] == "Sunday" || d["day"] == "Saturday") {
                        factor("Weekend")
                }
                else {
                        factor("Weekday")  
                }
        })
        partData
}
```

## What is mean total number of steps taken per day?
```{r}
totalNumberStepsPerDay <- function(data) {
        require("plyr")
        require("ggplot2")
        d <- ddply(data, c("date"),  function(d) {
                c(sum(d$steps, na.rm=TRUE))})
        names(d)[2] <- "sum"
        barplot(d$sum, xlab="Date", ylab="Total Steps" )
}
```

The following barchart shows the total number of steps taken per day
```{r echo=FALSE}
totalNumberStepsPerDay(getData())
```

## What is the mean and median of the number of steps taken per day
```{r}
reportMeanAndMedian <- function(data) {
        require("plyr")
        d1 <- ddply(data, c("date"),  function(d) {
                c(mean(d$steps, na.rm=TRUE))})
        d2 <- ddply(data, c("date"),  function(d) {
                c(median(d$steps, na.rm=TRUE))})
        d1 <- cbind(d1, d2$V1)
        names(d1)[2] <- "mean"
        names(d1)[3] <- "median"
        d1
}
```

The following table shows the mean and median of number of steps taken per day
```{r echo=FALSE}
reportMeanAndMedian(getData())
```

## What is the average daily activity pattern?
```{r}
averageDailyActivity <- function(data) {
        require("plyr")
        d <- ddply(data, c("interval"),  function(d) {
                c(mean(d$steps, na.rm=TRUE))})
        names(d)[2] <- "mean"
        plot(d$interval, d$mean, type="l", xlab="Intervals", ylab="Average Steps")
        print(paste("Interval with maximum average steps:",
                    d[which.max(d$mean), ]$interval))
}
```

The following chart shows the average number of steps taken per interval throughout a typical day
```{r echo=FALSE}
averageDailyActivity(getData())
```

## Inputing missing values
```{r}
inputMissingValues <- function(data) {
        #replace all the NAs with zero
        data[is.na(data)] <- 0
        data
}
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
plotWeekdayWeekend <- function(data) {
        require("plyr")
        require("lattice")
        
        dWeekday <- subset(data, data$weekday == "Weekday")
        dWeekend <- subset(data, data$weekday == "Weekend")

        d1 <- ddply(dWeekday, c("interval"),  function(d) {
                c(mean(d$steps, na.rm=TRUE))})
        d1$weekday <- factor("Weekday")
        
        d2 <- ddply(dWeekend, c("interval"),  function(d) {
                c(mean(d$steps, na.rm=TRUE))})
        d2$weekday <- factor("Weekend")
        
        plotData <- rbind(d1, d2)
        
        xyplot(plotData$V1~plotData$interval|plotData$weekday, 
                    main="Steps per Intervals", ylab="Steps",
                    xlab="Interval", type="l")
}
```

The following comparative charts show the difference in number of steps on a weekday vs. weekend
```{r echo=FALSE}
plotWeekdayWeekend(inputMissingValues(getData()))
```